Closing the gap between statistical and scientific workflows for improved inference in ecology

### 2025-07-24 revision/forecastflows_r1.tex
Changes to abstract

\begin{abstract}
% Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts. Datasets relevant for developing these models and forecasts have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare most current models for several reasons. First, the research community is split between two disconnected paradigms: either using data to fit simple, trend-like models with one or few parameters, or developing forecasting models that include complex, mechanistic, and/or black box submodels only indirectly informed by data. Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process. We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.
Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models. Datasets relevant for developing these models have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. 
%emw24Jul: Are you sure you want 'paradigm'? I love the word and the sentence but I think we mean more two 'groups' or 'spheres' -- I would switch and then carry the language you use to the next sentence. 
We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare current models for several reasons. First, the research community is split between two disconnected paradigms that prevent uniting a natural connection between trend estimation and forecasting. One research sphere focuses on using data to fit simple, trend-like models with few parameters, which is separated from research developing forecasting models, which often include complex, mechanistic sub-models only indirectly informed by data. Second, in both cases, models tend to be developed without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions throughout the entire modeling process. We propose a principled workflow that can address both of these limitations, by integrating statistical and scientific practices through clear steps, many of which rely on data simulation to inform decisions in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could improve our understanding of ecological systems and forecasts for them.

%emw24Jul: I pulled out a few of the best sentences that may help the abstract: 
% Implementing sustainable policies to mitigate these impacts is thus a global priority, but designing the best policies requires estimating and understanding biodiversity and ecosystem trends to date, alongside the skill to forecast future dynamics. 
% Towards this aim, we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.
% Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.

>>>>

\begin{abstract}
% Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts. Datasets relevant for developing these models and forecasts have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare most current models for several reasons. First, the research community is split between two disconnected paradigms: either using data to fit simple, trend-like models with one or few parameters, or developing forecasting models that include complex, mechanistic, and/or black box submodels only indirectly informed by data. Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process. We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.
Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models. Datasets relevant for developing these models have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. 
%emw24Jul: Are you sure you want 'paradigm'? I love the word and the sentence but I think we mean more two 'groups' or 'spheres' -- I would switch and then carry the language you use to the next sentence. 
We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare current models for several reasons. First, the research community is split between two disconnected paradigms that prevent uniting a natural connection between trend estimation and forecasting. One research sphere focuses on using data to fit simple, trend-like models with few parameters, which is separated from research developing forecasting models, which often include complex, mechanistic sub-models only indirectly informed by data. Second, in both cases, models tend to be developed without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions throughout the modeling process. To address these challenges we propose a workflow that integrates statistical and scientific practices through clear steps, many of which rely on data simulation to inform decisions in the process, and where forecasting a natural output. We show how this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could harmonize currently divided efforts at trend estimation and forecasting to better inform sustainable policies.

%emw24Jul: I pulled out a few of the best sentences that may help the abstract: I worked some in, see what you think. 
% Implementing sustainable policies to mitigate these impacts is thus a global priority, but designing the best policies requires estimating and understanding biodiversity and ecosystem trends to date, alongside the skill to forecast future dynamics. 
% Towards this aim, we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.
% Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.
